### 2010-03-30 papers/icfp-2010-desc/paper_desc.tex
Some suggestions to fix some typos

To model these data-types, we simply need to capture this grammar in
our dependently-typed setting. Here, the notion of sum-of-product
naturally translates into \emph{sigmas-of-sigmas}. However, beyond
this intuition, it should be clear that a direct encoding of
data-types through raw $\Sigma$-types is not a viable option. Indeed,
just as for finite sets, encoding throw away information when we
crucially need it. In the realm of data-types, things become even more
tougher, as one could legitimately expect an induction principle, for
instance. Anonymous in a sea of $\Sigma$s, our data-types have little
chance of survival. 
\note{Ridiculous tentative of poetry. I was really tired when I wrote that.}

@@

The key idea behind universe construction is our ability to make names
by defining new types. These names are called \emph{codes}. By
defining a set of codes, we somehow define the syntax of a
language. However, as such, a system of code is useless as it lacks a
semantics. Instead of equipping the universe of codes with some
computational behaviour, we pragmatically chose to \emph{interpret}
these codes back into the standard type theory. Hence, codes act only
as labels, while the type theory provides the computational
machinery. \note{Is that a good description?}

@@

To give some intuition on this universe of descriptions, we now turn
to some examples. For obvious pedagogical reasons, we will manually
build these descriptions. However, it should be clear that, in
practice, these definitions can be automatically constructed from an
Haskell-like $\data$ definition.

@@

So far, for convenience, we have taken as granted the existence of
$\Desc$, as presented in Figure~\ref{fig:desc_universe}. In
particular, we have considered its code, $\DSigma{\!}{\!}$,
$\DIndx{\!}$, and $\DUnit$ as type formers, extending the basic type
theory. Hence, in an implementation of this type theory, we need to
extended it with these constructors and their typing rules. Although
it makes no conceptual difference, the code of description should
rather be read as a specification. We promise the existence of such
objects, satisfying the typing rules. It will be the subject of
Section~\ref{sec:desc-levitate} to fulfil this promise, by actually
implementing the specification.

@@

So far, we have used our universe of descriptions to build
functors. \note{ Strictly positive types anyone? } To illustrate its
usage, we have implemented the pattern functors of natural numbers,
lists, and binary trees. However, the class of functors expressible
with descriptions enjoy another property: they all admit a
fix-point. Indeed, descriptions are interpreted into \emph{polynomial
  functors}~\cite{who?}\note{who is the relevant source for polynomial
  functors?}, as they are only composed of $\Sigma$-types, the unit
type, and variables.

@@

\[
\begin{array}{ll}
%%
\stk{
\begin{array}{@{}ll}
\All{\!}{\!}{\!}{\!} : & \PITEL{\V{D}}{\Desc}
                         \PITEL{\V{X}}{\Set}
                         \PITEL{\V{P}}{\V{X} \To \Set} \\
                       & \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                         \Set 
\end{array} \\
\All{\DId}{\V{X}}{\V{P}}{\V{x}} \mapsto \V{P}\: \V{x} \\
\All{(\DConst{\V{Z}})}{\V{X}}{\V{P}}{\V{z}} \mapsto 
    \Unit \\
\All{(\DProd{\V{D}}{\V{D'}})}{\V{X}}{\V{P}}{\pair{\V{d}}{\V{d'}}{}} \mapsto 
    \TIMES{\All{\V{D}}{\V{X}}{\V{P}}{\V{d}}}{\All{\V{D'}}{\V{X}}{\V{P}}{\V{d'}}} \\
\All{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} \mapsto 
    \All{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{b}} \\
\All{(\DPi{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{f}} \mapsto 
    \PI{\V{\V{s}}}{\V{S}} \All{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{(\V{f}\: \V{s})}
}
&
%%
\stk{
\begin{array}{@{}ll}
\all{\!}{\!}{\!}{\!}{\!} : & \PITEL{\V{D}}{\Desc}
                             \PITEL{\V{X}}{\Set}
                             \PITEL{\V{P}}{\V{X} \To \Set} \\
                           & \PITEL{\V{R}}{\PI{\V{x}}{\V{X}} \V{P}\: \V{x}}
                             \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                             \All{\V{D}}{\V{X}}{\V{P}}{\V{xs}} 
\end{array} \\
\all{\DId}{\V{X}}{\V{P}}{\V{R}}{\V{x}} = 
    \V{R}\: \V{x} \\
\all{(\DConst{\V{Z}})}{\V{X}}{\V{P}}{\V{R}}{\V{z}} = 
    \Void \\
\all{(\DProd{\V{D}}{\V{D'}})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{d}}{\V{d'}}{}} =
    \pair{\all{\V{D}}{\V{X}}{\V{P}}{\V{R}}{\V{d}}}{\all{\V{D'}}{\V{X}}{\V{P}}{\V{R}}{\V{d'}}}{} \\
\all{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{a}}{\V{b}}{}} = 
    \all{(\V{T}\: \V{a})}{\V{X}}{\V{P}}{\V{R}}{\V{b}} \\
\all{(\DPi{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{R}}{\V{f}} = 
    \LAM{\V{s}} \all{(\V{T}\: \V{s})}{\V{X}}{\V{P}}{\V{R}}{(\V{f}\: \V{s})}
\end{array}
}
\]

@@

Doing so, we seal data-type definition behind a convenient
abstraction. We take advantage of this abstraction barrier to tamper
the second issue. Namely, the developer would rather be presented a
name rather than a fix-point of a pattern functor. Surprisingly, it is
sufficient to \emph{label} the $\Mu{}$ constructor with the
user-defined name. When pretty-printing this object, we just expose
the label, instead of the definition in full.

Regaining the ability to write data-type constructor is simply a
matter of type propagation. We therefore extends the type-checking
framework of Figure~\ref{fig:type-checking} with the following
inference rule:
\note{ This is rather drafty. I should explain what $\vec{e}$ and $\widetilde{e'}$ are. }

\[
\Rule{\forall \V{e_i} \in \V{\vec{e}}, \propag{\push{\V{e_i}}
                                            {\descop{\V{f}\: \V{i}}{(\Mu{(\DSigma{(\Enum\: \V{b})}{\V{f}})})}}}
                                      {\V{e'_i}}}
     {\propag{\push{\V{C}\: \V{\vec{e}}}{\Mu{(\DSigma{(\Enum\: \V{b})}{\V{f}})}}}
             {\Con{(\etag{C}, \V{\widetilde{e'}})}}}\;\etag{t} \in \V{b}
\]

Hence, in the high-level expression language, the developer writes a
familiar type constructors, applied to a telescope of arguments. In
the type propagation setting, this expression will be pushed to a
$\Mu{}$ type of a tagged description. It is then straightforward to
look-up the tag and turn the constructor form into the appropriate
code.

>>>>

To model these data-types, we simply need to capture this grammar in
our dependently-typed setting. Here, the notion of sum-of-product
naturally translates into \emph{sigmas-of-sigmas}. However, beyond
this intuition, it should be clear that a direct encoding of
data-types through raw $\Sigma$-types is not a viable option. Indeed,
just as for finite sets, encoding throws away information when we
crucially need it. In the realm of data-types, things become even
tougher, as one could legitimately expect an induction principle, for
instance. Anonymous in a sea of $\Sigma$s, our data-types have little
chance of survival. 
\note{Ridiculous tentative of poetry. I was really tired when I wrote that.}

@@

The key idea behind universe construction is our ability to make names
by defining new types. These names are called \emph{codes}. By
defining a set of codes, we somehow define the syntax of a
language. However, as such, a system of codes is useless as it lacks a
semantics. Instead of equipping the universe of codes with some
computational behaviour, we pragmatically chose to \emph{interpret}
these codes back into the standard type theory. Hence, codes act only
as labels, while the type theory provides the computational
machinery. \note{Is that a good description?}

@@

To give some intuition on this universe of descriptions, we now turn
to some examples. For obvious pedagogical reasons, we will manually
build these descriptions. However, it should be clear that, in
practice, these definitions can be automatically constructed from a
Haskell-like $\data$ definition.

@@

So far, for convenience, we have taken as granted the existence of
$\Desc$, as presented in Figure~\ref{fig:desc_universe}. In
particular, we have considered its codes, $\DSigma{\!}{\!}$,
$\DIndx{\!}$, and $\DUnit$ as type formers, extending the basic type
theory. Hence, in an implementation of this type theory, we need to
extended it with these constructors and their typing rules. Although
it makes no conceptual difference, the code of description should
rather be read as a specification. We promise the existence of such
objects, satisfying the typing rules. It will be the subject of
Section~\ref{sec:desc-levitate} to fulfil this promise, by actually
implementing the specification.

@@

So far, we have used our universe of descriptions to build
functors. \note{ Strictly positive types anyone? } To illustrate its
usage, we have implemented the pattern functors of natural numbers,
lists, and binary trees. However, the class of functors expressible
with descriptions enjoys another property: they all admit a
fix-point. Indeed, descriptions are interpreted into \emph{polynomial
  functors}~\cite{who?}\note{who is the relevant source for polynomial
  functors?}, as they are only composed of $\Sigma$-types, the unit
type, and variables.

@@

\[
\begin{array}{ll}
%%
\stk{
\begin{array}{@{}ll}
\All{\!}{\!}{\!}{\!} : & \PITEL{\V{D}}{\Desc}
                         \PITEL{\V{X}}{\Set}
                         \PITEL{\V{P}}{\V{X} \To \Set} \\
                       & \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                         \Set 
\end{array} \\
\All{\DUnit}{\V{X}}{\V{P}}{\Void} = 
    \Unit \\
\All{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} = 
    \All{(\V{T}\: a)}{\V{X}}{\V{P}}{\V{b}} \\
\All{(\DIndx{\V{D}})}{\V{X}}{\V{P}}{\pair{\V{a}}{\V{b}}{}} =
    \TIMES{\V{P}\: \V{a}}{\All{\V{D}}{\V{X}}{\V{P}}{\V{b}}} \\
\All{(\DHindx{\V{H}}{\V{D}})}{\V{X}}{\V{P}}{\pair{\V{f}}{\V{b}}{}} =
    \TIMES{(\PI{\V{i}}{\V{H}} \V{P}\: (\V{f}\: \V{i}))}
          {\All{\V{D}}{\V{X}}{\V{P}}{\V{b}}}
}
&
%%
\stk{
\begin{array}{@{}ll}
\all{\!}{\!}{\!}{\!}{\!} : & \PITEL{\V{D}}{\Desc}
                             \PITEL{\V{X}}{\Set}
                             \PITEL{\V{P}}{\V{X} \To \Set} \\
                           & \PITEL{\V{R}}{\PI{\V{x}}{\V{X}} \V{P}\: \V{x}}
                             \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                             \All{\V{D}}{\V{X}}{\V{P}}{\V{xs}} 
\end{array} \\
\all{\DUnit}{\V{X}}{\V{P}}{\V{R}}{\Void} =
    \Void \\
\all{(\DSigma{\V{S}}{\V{T}})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{a}}{\V{b}}{}} =
    \all{(\V{T}\: a)}{\V{X}}{\V{P}}{\V{R}}{\V{b}} \\
\all{(\DIndx{D})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{a}}{\V{b}}{}} =
    \pair{\V{R}\: \V{a}}
         {\all{\V{D}}{\V{X}}{\V{P}}{\V{R}}{\V{b}}}{} \\
\all{(\DHindx{H}{D})}{\V{X}}{\V{P}}{\V{R}}{\pair{\V{f}}{\V{b}}{}} =
    \pair{\LAM{i} \V{R}\: (\V{f}\: \V{i})}
         {\all{\V{D}}{\V{X}}{\V{P}}{\V{R}}{b}}{}
\end{array}
}
\]

@@

Doing so, we seal data-type definition behind a convenient
abstraction. We take advantage of this abstraction barrier to address
the second issue. Namely, the developer would rather be presented a
name than a fix-point of a pattern functor. Surprisingly, it is
sufficient to \emph{label} the $\Mu{}$ constructor with the
user-defined name. When pretty-printing this object, we just expose
the label, instead of the definition in full.

Regaining the ability to write data-type constructors is simply a
matter of type propagation. We therefore extend the type-checking
framework of Figure~\ref{fig:type-checking} with the following
inference rule:

\[
\Rule{\propag{\push{\widetilde{\V{e}}}
                   {\descop{\V{f}\: \V{C}}{(\Mu{(\DSigma{(\Enum\: \V{b})}{\V{f}})})}}}
            {\widetilde{\V{e}}'}}
     {\propag{\push{\V{C}\: \vec{\V{e}}}
                   {\Mu{(\DSigma{(\Enum\: \V{b})}{\V{f}})}}}
             {\Con{(\etag{\V{C}}, \widetilde{\V{e}}')}}}\;\etag{\V{C}} \in \V{b}
\]

Where $\vec{\V{e}}$ denotes a telescope of arguments applied to the
constructor $\V{C}$. For the sake of readability, we write function
application in the small function space as $\V{f}\: \V{C}$. This
corresponds to a call to $\F{switch}$. Also, we note
$\widetilde{\V{e}}$ the right-nested tuple terminated by $\Void$
corresponding to the curried telescope $\vec{\V{e}}$.

Hence, in the high-level expression language, the developer writes a
familiar type constructor, applied to a telescope of arguments. In
the type propagation setting, this expression will be pushed to a
$\Mu{}$ type of a tagged description. It is then straightforward to
look-up the tag and turn the constructor form into the appropriate
code.

### 2010-03-30 papers/icfp-2010-desc/paper_desc_levitation.tex
Some suggestions to fix some typos

In Section~\ref{sec:finite-sets}, we have given a specification of
finite sets. It consists in the objects presented in
Figure~\ref{fig:typing-finite-set}. We are going to implement the
$\EnumU$ type former and its constructors. Let us recall their
specification:

@@

The names $\NilE$ and $\ConsE{\!}{\!}$ are absolutely not a
coincidence. $\EnumU$ is indeed isomorphic to a list of $\UId$s. In
Section~\ref{sec:desc-examples}, we have shown how to build lists in
the universe of description. Therefore, the actual implementation of
$\EnumU$ boils down to the following, trivial definition:

@@

Let us examine the consequences of our act. First of all, we discover
that the type theory does not actually need to be extended with the
type former $\EnumU$, the constructors $\NilE$ and $\ConsE$. We have
just defined $\EnumU$ above, by fix-point over the pattern functor denoted
by $\ListD$. $\NilE$ and $\ConsE$ are simply the $\ListNil$ and $\ListCons$
constructors of lists.

@@

Apart from these changes, we are left with implementing the other
components of finite sets. This consists in $\EnumT$, $\Ze$, $\Su$, as
well as the $\F{switch}$ operator. Note that the actual implementation
of $\EnumU$ does not influence our implementation of $\EnumT$: be they
hard-coded or codes in the $\Desc$ universe, the $\EnumU$ objects just
behave similarly. We have witnessed this effect when carrying this
operation in Epigram, moving from an hard-coded presentation to a
self-hosted one. Absolutely no change to the $\EnumT$ objects was
required.

@@

In this section, we have replaced a low-level presentation of finite
sets by a self-hosted one, expressed in the universe of
descriptions. However, formally, the content of the type theory
remains unchanged: objects which were present before the modification
are still there. Conversely, we have not introduced any spurious
object.

@@

Our first tentative attempt is the following:

@@

In order to capture a notion of higher-order induction, we need to
extend our universe of descriptions. This consists in adding a code
$\DHindx{\!}{\!}$ that takes an indexing set $\V{H}$. The code and its
interpretation are presented in
Figure~\ref{fig:hindx_desc}. Intuitively, $\DHindx{\!}{\!}$ use the
elements of $H$ to index as many recursive arguments.

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \ldots          & \:\:\ldots \\
    \DHindx{\!}{\!} & \PI{\V{H}}{\Set} \Desc \To \Desc
\end{array}
}
\vspace{0.2in}
\\
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\ldots                        &  \ldots \\
\descop{\DHindx{\V{H}}{\V{D}}}{\V{X}}     &  \TIMES{(\V{H} \To \V{X})}{\descop{\V{D}}{\V{X}}}
\end{array}
}
\end{array}
}\]

@@

Note that $\DIndx{\!}$ is isomorphic to $\DHindx{\Unit}{\!}$. However,
we tolerate this duplication and do not replace $\DIndx{\!}$ by the
more general $\DHindx{\!}{\!}$. Indeed, unlike its counterpart,
$\DIndx{\!}$ is first-order. Therefore, while both codes are
\emph{extensionally} the same, in most practical implementation they
would be dealt with rather differently. While definitional equality
can cope with first order objects, the functional presentation
introduced by $\DHindx{\!}{\!}$ is unlikely to be amenable to a purely
definitional treatment.

@@

In a first glance, we have achieved our goal. We have described the
codes of the universe of description. Taking the fix-point of this
object gives us $\Desc$, up to isomorphism. We have implemented our
code of data-types as an object levitating inside itself. However,
this levitation operation, just as any magic trick, relies on an
invisible cable. Let us reveal it, hence finishing our implementation.

@@

Our reader might be slightly confused to learn that the trick is
visible in the definition of $\DescD$
(Fig.~\ref{fig:desc-levitate}). Or rather, it is made invisible by
careful usage of type propagation. Indeed, let us try to elaborate
this term down to the low-level type theory. The $\EnumT{[ \ldots ]}$
construct elaborates to a finite set $e$ in $\EnumU$, inhabited by the
codes. Then, we can type-check the case definition, between square
brackets $[ \ldots ]$. This term $\V{f}$ is pushed into the type
$\EnumT{\V{e}} \To \Desc$. This corresponds to a finite function
definition, as formalised in Figure~\ref{fig:type-checking}.

@@

We are left with implementing filling the hole. Recall that we have
\(\Bhab{\V{xs}}{\descop{\V{D}}{\Mu{\V{D}}}}\) and
\(\Bhab{\V{hs}}{\All{\V{D}}{(\Mu{\V{D}})}{(\LAM{\_} \V{T})}{\V{xs}}}\) at hand. Our goal
is to make an element of \(\descop{\V{D}}{\V{T}}\). Intuitively, $\V{xs}$ is of
the right shape, but its sub-elements are of the wrong type. On the
other hand, for each sub-element of $\V{xs}$, $\V{hs}$ gives us the
corresponding element in $\V{T}$. Hence the name ``induction
hypotheses''. Therefore, to construct an element of \(\descop{\V{D}}{\V{T}}\),
we replace the recursive components of \(\V{xs}\) by their counterpart in
\(\V{hs}\):

\[\stk{
\F{replace} : \stk{\PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X},\V{Y}}{\Set}\\
                   \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                   \All{\V{D}}{\V{X}}{(\LAM{\_}\V{Y})}{\V{xs}} \To
                   \descop{\V{D}}{\V{Y}}} \\
\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void          \mapsto \Void \\
\F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{xs}}{}\: \V{ys}             \mapsto
    \pair{\V{s}}{\F{replace}\: {\V{D}~\V{s}}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}                                 \\
\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{xs}}{}\: \pair{\V{y}}{\V{ys}}{} \mapsto
    \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}
}\]

@@

@@

With this example, we have shown how we can derive a generic
operation, the catamorphism, from a pre-existing generic operation,
the induction principle. This has been made possible by our ability to
manipulate descriptions as first-class objects: the catamorphism is,
basically, a function mapping a $\Desc$ to a data-type specific
operation.

@@

\[
\stk{
\data \FMFreeMonad : \PITEL{\V{F}}{\Set \To \Set} 
                     \PI{\V{X}}{\Set} 
                     \Set 
\where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FMVar           & \V{X} \To \FMFreeMonad\: \V{F}\: \V{X}                            \\
    \FMComposite     & \V{F} (\FMFreeMonad\: \V{F}\: \V{X}) \To \FMFreeMonad\: \V{F}\: \V{X}    
\end{array}
}
\]

@@

We simply add a constructor, $\DVar{\!}$, and define its argument to
be a $\DSigma{\V{X}}{\DUnit}$, that is an element of $\V{X}$. We keep
$\V{E}$ and $\V{D}$ as they were, hence leaving the functor
unchanged. Unfolding the interpretation of this definition, we
convince ourself that this corresponds to the functor
$\FMFreeMonadD$. The fix-point operation ties the knot and gives us
the full-blown free monad construction.

@@

Of course, we must equip the resulting data-types with operations
delivering a monadic interface. As usual, \(\LAM{\x}\DVar{\x}\)
performs the r\^ole of \return, embedding variables into terms. The
\bind operation corresponds to \emph{substitution}. We will now
implement it, as a generic function.

@@

Our task is therefore to implement the algebra of the
catamorphism. Intuitively, its role is to catch appearances of
$\DVar{\V{x}}$ and replace them by $\V{\sigma}\: \V{x}$. Otherwise, it should simply
appeal to the induction hypothesis. This corresponds to the following
definition:

@@

Filling the sub-goal with $\apply\: \V{D}\: \V{X}\: \V{Y}\: \V{\sigma}$
completes the implementation. To sum up, we have implemented the free
monad construction for an arbitrary tagged description. This gives the
developer the ability, for any data-type, to extend it with a notion
of variable. Then, we have equipped this structure with the
corresponding monadic operation, \bind and \return. Whereas the
\return is a trivial variable former, the \bind operation corresponds
to the substitution of variables by terms. We have shown how to
implement it, for any free monad.

@@

With the free monad construction, we have seen two kind of generic
operations. Firstly, we have derived a new data-structure from another
one: we make the free monad from its underlying functor. To do so, we
crucially rely on the fact that data-types are nothing but codes. We
are therefore entitled to modify this code and, in this case, extend
it. Extending a data-type might give rise to a more structured
object, as was the case here.

So, secondly, we have equipped this new data-type with its inherent
structure: the \bind and \return operations. We have been able to
build them as generic functions. Here, we rely on our ability to
compute over descriptions.

>>>>

In Section~\ref{sec:finite-sets}, we have given a specification of
finite sets. It consists of the objects presented in
Figure~\ref{fig:typing-finite-set}. We are going to implement the
$\EnumU$ type former and its constructors. Let us recall their
specification:

@@

The names $\NilE$ and $\ConsE{\!}{\!}$ are not a coincidence. $\EnumU$
is indeed isomorphic to a list of $\UId$s. In
Section~\ref{sec:desc-examples}, we have shown how to build lists in
the universe of description. Therefore, the actual implementation of
$\EnumU$ boils down to the following, trivial definition:

@@

Let us examine the consequences. First of all, we discover that the
type theory does not actually need to be extended with the type former
$\EnumU$, the constructors $\NilE$ and $\ConsE$. We have just defined
$\EnumU$ above, by fix-point over the pattern functor denoted by
$\ListD$. $\NilE$ and $\ConsE$ are simply the $\ListNil$ and
$\ListCons$ constructors of lists.

@@

Apart from these changes, we are left with implementing the other
components of finite sets. This consists in $\EnumT$, $\Ze$, $\Su$, as
well as the $\F{switch}$ operator. Note that the actual implementation
of $\EnumU$ does not influence our implementation of $\EnumT$: be they
hard-coded or codes in the $\Desc$ universe, the $\EnumU$ objects
behave the same. We have witnessed this effect when carrying the
operation in Epigram, moving from an hard-coded presentation to a
self-hosted one. Absolutely no change to the $\EnumT$ objects was
required.

@@

In this section, we have replaced a low-level presentation of finite
sets by a self-hosted one, expressed in the universe of
descriptions. However, formally, the content of the type theory
remains unchanged: objects that were present before the modification
are still there. Conversely, we have not introduced any spurious
object.

@@

Our first attempt is the following:

@@

In order to capture a notion of higher-order induction, we need to
extend our universe of descriptions. This consists in adding a code
$\DHindx{\!}{\!}$ that takes an indexing set $\V{H}$. The code and its
interpretation are presented in
Figure~\ref{fig:hindx_desc}. Intuitively, $\DHindx{\!}{\!}$ uses the
elements of $H$ to index as many recursive arguments.

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DHindx{\!}{\!} & \PI{\V{H}}{\Set} \Desc \To \Desc \\
    \ldots          & \:\:\ldots
\end{array}
}
\vspace{0.2in}
\\
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}l@{\:=\:\:}ll}
\descop{\DHindx{\V{H}}{\V{D}}}{\V{X}}     &  \TIMES{(\V{H} \To \V{X})}{\descop{\V{D}}{\V{X}}} \\
\ldots                                    &  \ldots 
\end{array}
}
\end{array}
}\]

@@

Note that $\DIndx{\!}$ is isomorphic to $\DHindx{\Unit}{\!}$. However,
we tolerate this duplication and do not replace $\DIndx{\!}$ by the
more general $\DHindx{\!}{\!}$. Indeed, unlike its counterpart,
$\DIndx{\!}$ is first-order. Therefore, while both codes are
\emph{extensionally} the same, in most practical implementations they
would be dealt with rather differently. While definitional equality
can cope with first order objects, the functional presentation
introduced by $\DHindx{\!}{}$ is unlikely to be amenable to a purely
definitional treatment.

@@

At a first glance, we have achieved our goal. We have described the
codes of the universe of description. Taking the fix-point of this
object gives us $\Desc$, up to isomorphism. We have implemented our
code of data-types as an object levitating inside itself. However,
this levitation operation, just as any magic trick, relies on an
invisible cable. Let us reveal it, hence finishing our implementation.

@@

Our reader might be slightly confused to learn that the trick is
visible in the definition of $\DescD$
(Fig.~\ref{fig:desc-levitate}). Or rather, it is made invisible by
careful usage of type propagation. Indeed, let us try to elaborate
this term down to the low-level type theory. The $\EnumT{[ \ldots ]}$
construct elaborates to a finite set $e$ in $\EnumU$, inhabited by the
codes. Then, we can type-check the case definition, between square
brackets $[ \ldots ]$. This term $\V{f}$ is pushed into the type
$\EnumT{\V{e}} \To \Mu{\DescD}$. This corresponds to a finite function
definition, as formalised in Figure~\ref{fig:type-checking}.

@@

We are left with filling the hole. Recall that we have
\(\Bhab{\V{xs}}{\descop{\V{D}}{\Mu{\V{D}}}}\) and
\(\Bhab{\V{hs}}{\All{\V{D}}{(\Mu{\V{D}})}{(\LAM{\_} \V{T})}{\V{xs}}}\)
at hand. Our goal is to make an element of
\(\descop{\V{D}}{\V{T}}\). Intuitively, $\V{xs}$ is of the right
shape, but its sub-elements are of the wrong type. On the other hand,
for each sub-element of $\V{xs}$, $\V{hs}$ gives us the corresponding
element in $\V{T}$. Hence the name ``induction
hypotheses''. Therefore, to construct an element of
\(\descop{\V{D}}{\V{T}}\), we replace the recursive components of
\(\V{xs}\) by their counterpart in \(\V{hs}\):

\[\stk{
\F{replace} : \stk{\PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X},\V{Y}}{\Set}\\
                   \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                   \All{\V{D}}{\V{X}}{(\LAM{\_}\V{Y})}{\V{xs}} \To
                   \descop{\V{D}}{\V{Y}}} \\
\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void          \mapsto 
    \Void                                                                                                         \\
\F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{xs}}{}\: \V{ys}             \mapsto
    \pair{\V{s}}{\F{replace}\: {\V{D}~\V{s}}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}                                 \\
\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{xs}}{}\: \pair{\V{y}}{\V{ys}}{} \mapsto       \\
\qquad  \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}                                     \\
\F{replace}\: (\DHindx{\V{H}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{f}}{\V{xs}}{}\: \pair{\V{g}}{\V{ys}}{} \mapsto   \\
\qquad  \pair{\V{g}}{\F{replace}\: \V{D}\{ \V{X}\: \V{Y}\: \V{g}\: \V{ys}}{}
}\]

@@

The astute reader will have been struck by the type of $\F{replace}$:
it is \emph{almost} the morphism part -- sometimes called \emph{map}
-- of the functor $\V{D}$ from $\V{X}$ to $\V{Y}$ in $\Set$. Just as
the $\F{induction}$ is the dependent version of $\cata$, $\F{replace}$
is the dependent version of the map, which uses the induction
hypotheses. For space reason, we will not present the non-dependent
map. It can be found in the Agda model.

@@

With this example, we have shown how we can derive a generic
operation, the catamorphism, from a pre-existing generic operation,
the induction principle. This has been made possible by our ability to
manipulate descriptions as first-class objects: the catamorphism is,
basically, a function mapping a $\Desc$ to a data-type specific
operation. This programming style is reminiscent to polytypic
languages, such as PolyP~\cite{jansson:polyp}.

@@

\[
\stk{
\data \FMFreeMonad\: \PITEL{\V{F}}{\Set \To \Set} 
                     \PITEL{\V{X}}{\Set} :
                     \Set 
\where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FMVar           & \V{X} \To \FMFreeMonad\: \V{F}\: \V{X}                            \\
    \FMComposite     & \V{F} (\FMFreeMonad\: \V{F}\: \V{X}) \To \FMFreeMonad\: \V{F}\: \V{X}    
\end{array}
}
\]

@@

We simply add a constructor, $\DVar{\!}$, and define its argument to
be a $\DSigma{\V{X}}{\DUnit}$, that is an element of $\V{X}$. We keep
$\V{E}$ and $\V{D}$ as they were, hence leaving the functor
unchanged. Unfolding the interpretation of this definition, we
convince ourselves that this corresponds to the functor
$\FMFreeMonadD$. The fix-point operation ties the knot and gives us
the full-blown free monad construction.

@@

Of course, we must equip the resulting data-types with operations
delivering a monadic interface. As expected, \(\LAM{\x}\DVar{\x}\)
plays the r\^ole of \return, embedding variables into terms. The
\bind\ operation corresponds to \emph{substitution}. We will now
implement it, as a generic function.

@@

Our task is therefore to implement the algebra of the
catamorphism. Intuitively, its role is to catch appearances of
$\DVar{\V{x}}$ and replace them by $\V{\sigma}\: \V{x}$. This
corresponds to the following definition:

@@

Filling the sub-goal with $\apply\: \V{D}\: \V{X}\: \V{Y}\:
\V{\sigma}$ completes the implementation. To sum up, we have
implemented the free monad construction for an arbitrary tagged
description. This gives the developer the ability, for any data-type,
to extend it with a notion of variable. Then, we have equipped this
structure with the corresponding monadic operation, \bind\ and
\return. Whereas the \return\ is a trivial variable former, the
\bind\ operation corresponds to the substitution of variables by
terms. We have shown how to implement it, for any free monad. This
programming style is reminiscent to the type-indexed data
types~\cite{hinze:generic-haskell}, as implemented in Generic Haskell:
from data-types, we built new data-types and equip them with some
structure.

@@

With the free monad construction, we have seen two kind of generic
operations. Firstly, we have derived a new data-structure from another
one: we make the free monad from its underlying functor. To do so, we
crucially rely on the fact that data-types are nothing but codes. We
are therefore entitled to modify this code and, in this case, extend
it. Extending a data-type might give rise to a more structured object,
as was the case here.  So, secondly, we have equipped this new
data-type with its inherent structure: the \bind\ and
\return\ operations. We have been able to build them as generic
functions.
